---
title: "Area 10"
author: "Ty Garber"
date: "2023-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
tf <- pssp::tf_pull('10', '2023-07-13', '2023-08-03')
```

Calulate running test fishing counts by day
```{r, running_counts}
# running counts
tf_running_sum_counts <- tf %>%
  group_by(survey_datetime) %>%
  nest() %>%
  mutate(
    lm = map(data, pssp::tf_lm_summary)
  ) %>%
  unnest(lm) %>%
  select(-data) %>%
  ungroup() %>%
  mutate(
    across(where(is.integer), \(x) replace_na(x, 0)),
    across(where(is.integer), \(x) cumsum(x))
  )

tf_running_sum_counts
```

Calculate running percentages by day
```{r running_percentage}
tf_running_sum_percent <- tf_running_sum_counts %>%
  mutate(
    across(where(is.integer),\(x) x / (Legal_AD + Legal_UM + Sublegal_AD + Sublegal_UM))
  )
tf_running_sum_percent
```

Total encounters in the fishery are calculated by dividing the legal mark proportion
in the test fishery, VTRs or dockside encounters and applying the 13% bias correction.
  
$$
Total Encounters = \frac{Est. AD}{.87}\div LM\%
$$
The relationship between legal mark percentage and total encounters is not linear, with the number of encouners being estimated exponentially growing the lower the legal mark percentage is.

```{r}
tibble(
  legal_mark = seq(0.01, 1, by = .01),
  encounters_per_ret_ad = 1 / legal_mark / .87
) %>%
  ggplot(aes(legal_mark, encounters_per_ret_ad, group=1)) +
    geom_line() +
  labs(y = 'Estimated Encounters',
       x = 'Legal Mark Proportion',
       subtitle = 'Relationship Between Legal Mark Proportion and Estimated Encounters') +
  scale_x_continuous(labels = scales::percent)
```

To account for this great expansion when the legal mark proportion is low are great number of samples 
are needed to reach a point where the coefficient of variation meets the agreed-to criteria of 20%

```{r}
tibble(
  legal_mark = seq(0.01, 1, by = .01),
  sample_size = ((legal_mark - legal_mark^2) / (.2 * legal_mark)^2) + 1
) %>%
  ggplot(aes(legal_mark, sample_size, group=1)) +
    geom_line() +
  labs(y = 'Required Sample Size',
       x = 'Legal Mark Proportion',
       subtitle = 'Relationship Between Legal Mark Proportion and Sample Size need to achieve 20%') +
  scale_x_continuous(labels = scales::percent)
```

Legal mark percentages in the Area 10 Summer Chinook fishery have remained low through much of the fishery, 
only in recent test fishing coming up to a point where sample sizes have met the confidence of the 20% coefficient of variation.

```{r}
tf_running_sum_percent %>%
  ggplot(aes(survey_datetime, Legal_AD, group =1)) +
  geom_line() + 
  labs(
    subtitle = 'Legal Mark Percentage in MA10 Test Fishery',
    x = 'Date',
    y = 'Legal Mark Percentage'
  ) + 
  scale_y_continuous(labels = scales::percent)
```

Coefficient of variation over time. <span style="color:red"> Red Line </span> indicates 20% sample size goal.
```{r}
# coefficient of variation over time
tf_running_sum_percent %>%
  inner_join(tf_running_sum_counts, by = 'survey_datetime', suffix=c('_percent', '_count')) %>%
  ggplot(
    aes(
      survey_datetime,
      sqrt(((1 - Legal_AD_percent) * (Legal_AD_percent)) / 
        ((Legal_AD_count + Legal_UM_count + Sublegal_AD_count + Sublegal_UM_count) - 1))
      / Legal_AD_percent
    , group=1)
  ) + geom_line() +
  geom_hline(yintercept = .2, color = 'red') +
  labs(
    x = 'Date',
    y = 'CV% of Legal Mark Percentage'
  ) + 
  scale_y_continuous(labels = scales::percent)
```



Unlike the an M1 estimate, M2 estimates of encounters are independent from estimates of catch. Each test fishing survey will effect the estimate of encounters. Below are estimates of sublegal quota utilization vs FRAM predictions as the fishery progressed.
```{r}
catch <- tribble(
  ~date, ~est_ad_ret,
  '2023-07-13', 93,
  '2023-07-14', 66,
  '2023-07-15', 135,
  '2023-07-16', 124,
  '2023-07-20', 139,
  '2023-07-21', 53,
  '2023-07-22', 45,
  '2023-07-23', 85,
  '2023-07-27', 287,
  '2023-07-28', 88,
  '2023-07-29', 112,
  '2023-07-30', 126,
  '2023-08-03', 241
) %>% mutate(across(date, as.Date))

cv <- tf_running_sum_percent %>%
  inner_join(tf_running_sum_counts, by = 'survey_datetime', suffix=c('_percent', '_count')) %>%
  mutate(
    cv =  sqrt(((1 - Legal_AD_percent) * (Legal_AD_percent)) / 
        ((Legal_AD_count + Legal_UM_count + Sublegal_AD_count + Sublegal_UM_count) - 1))
      / Legal_AD_percent
  ) %>%
  select(survey_datetime, cv)

tf_running_sum_percent %>%
  inner_join(cv) %>%
  full_join(catch, by = c('survey_datetime' = 'date')) %>%
  arrange(survey_datetime) %>%
  fill(Legal_AD, Legal_UM, Sublegal_AD, Sublegal_UM,cv) %>%
  mutate(
    across(est_ad_ret, \(x) cumsum(replace_na(x, 0))),
    est_sublegal = (est_ad_ret / Legal_AD / .87) * (Sublegal_UM + Sublegal_AD),
    quota_percent = est_sublegal / 7748
  ) %>%
  pivot_longer(c(quota_percent, cv)) %>%
  select(survey_datetime, name, value) %>%
  ggplot(aes(survey_datetime, value, color=name, group=name)) +
  geom_line() +
  geom_hline(yintercept = 1, color='tomato4')+
  labs(x = 'Date',
       y = NULL,
       subtitle = 'MA10 Sublegal Quota vs FRAM as Fishery Progressed') +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.3)) +
  scale_color_manual(
    labels = c('CV%', 'Sublegal Quota %'),
    values = c('seagreen', 'midnightblue')
  ) +
  theme(legend.title = element_blank())

```


Legal Mark proportions encountered in the test fishery drastically changed as AD Chinook entered MA10 7/31-8/3. This has a large effect on the number of sublegals estimated for the fishery, falling by about half, from 7 sublegal encounters per retained AD chinook to around 3.5.

```{r}
tf_running_sum_percent %>%
  ggplot(aes(survey_datetime, (1 / Legal_AD / .87) * (Sublegal_AD + Sublegal_UM), group = 1)) +
           geom_line() +
  scale_y_continuous(breaks = seq(3,17, by = 1)) +
  labs(
    subtitle = 'Area 10 Estimated Encounters via Test Fishing',
    x = 'Date',
    y = 'Sublegal Encounters per AD Retained Chinook'
    )
```

This rise in legal mark percentage and drop of sublegal encounters coincides with the historic run
timing of legal marked Chinook in MA10. Here is the run timing as displaying through sport CPUE.

```{r}
ds <- pssp::ds_salmon_summary('10', '2007-01-01', '2023-08-31')

ds %>%
  filter(
    between(strftime(survey_datetime, '%m%d'), '0713', '0831')) %>%
  group_by(survey_datetime, month_day = strftime(survey_datetime, '%m%d')) %>%
  summarize(across(c(anglers, chinook_ad_k), \(x) sum(x, na.rm = T)), .groups='drop') %>%
  mutate(
    cpue = chinook_ad_k / anglers
  ) %>%
  ggplot(aes(lubridate::mdy(paste0(month_day, 2023)), cpue)) +
  geom_smooth(method='gam') +
  labs(
    y = 'Chinook AD / Angler',
    x = 'Date',
    subtitle = 'Historic Run Timing as Measure by Sport CPUE in MA10'
  ) +
  scale_x_date(breaks = '1 day') +
  theme(axis.text.x = element_text(angle = 90, vjust=.5, hjust=1))
```

